Viral and Bacterial Communities of Colorectal Cancer

### 2017-04-11 doc/manuscript.Rmd
Updates and edits.

```{r loadfiles, echo=FALSE}
sigfig = 3
alphadf <- read.delim(file = "../rtables/diversity.tsv", header = TRUE, sep = "\t")
twoauc <- read.delim(file = "../rtables/twoclass-auc.tsv", header = TRUE, sep = "\t")
twoss <- read.delim(file = "../rtables/twoclass-auc-statsig.tsv", header = TRUE, sep = "\t")
threeauc <- read.delim(file = "../rtables/threeclass-auc.tsv", header = TRUE, sep = "\t")
threesig <- read.delim(file = "../rtables/threeclass-sig.tsv", header = TRUE, sep = "\t")
idcount <- read.delim(file = "../rtables/idcount.tsv", header = FALSE, sep = "\t")
phagecount <- read.delim(file = "../rtables/phagecount.tsv", header = FALSE, sep = "\t")
corstats <- read.delim(file = "../rtables/cor-stats.tsv", header = TRUE, sep = "\t")
```

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer). Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases suggest that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans have died from the disease [@Siegel:2014jo]. An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those changes might impact cancer progression and severity. We also aimed to evaluate their potential for use as diagnostic biomarkers. The implications of this study are threefold. *First*, this work will support a biological role for the virome in colorectal cancer development and suggest that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Cohort Design, Sample Collection, and Processing
Our study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**. Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The other half of the stool samples were purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that were encapsulated.

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OVU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics. We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Similar to what was previously seen when analyzing the 16S rRNA gene sequences from the same samples [@Zackular:2014fba], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.

@@

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model yielded a statistically significant but minor performance improvement beyond the viral (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that the combined features from the virus and bacterial communities improve our ability to classify stool as belonging to individuals with cancerous colons.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed more poorly than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**. Although the features in the viral model were also of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven by a few OVUs, suggesting a role for the viruses in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome. This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Shifted Phage Influence Across Cancer Stages
Because our cohort included people with healthy, adenomatous, and cancerous colons we were able to evaluate shifts in influential phage identities during cancer progression. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. Not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.

## Bacteriophage Dominance in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

@@

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network. We investigated the infectious capabilities and potential host ranges of all phage OVUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**). The predicted interactions were then used to identify community hub phages. This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OVU's connection to the rest of the network, and compared the centrality to the importance of each OVU in the colorectal cancer classification model. The phages with high centrality values were defined as community hubs. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

@@

A notable observation from our analysis was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences. This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing. We found that there were six bacterial OTUs that drove the performance of the 16S rRNA gene classification model, and these OTUs were all sparsely present and lowly abundant. Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model. The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application. There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA gene model [@Zeller:2014ix]. Thus the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples. Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.

@@

## Analysis Source Code & Availability
All associated source code is available at the following GitHub repository:

@@

## Virus Metagenomic Library Preparation & Sequencing
Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz]. Briefly, an aliquot of stool (~0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension. The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22-µm filter to remove smaller contaminants. The filtered supernatant was treated with chloroform to lyse contaminating cells including bacteria, human, fungi, etc. The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase. The DNA was extracted from the purified VLPs using the Wizard PCR Purification Preparation Kit (Promega). Disease classes were staggered across purification runs to prevent run variation as a confounding factor. Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol. The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution. The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz]. The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.

>>>>

```{r loadfiles, echo=FALSE}
psigfig = 1
asigfig = 3
sigfig = 3

alphadf <- read.delim(file = "../rtables/diversity.tsv", header = TRUE, sep = "\t")
twoauc <- read.delim(file = "../rtables/twoclass-auc.tsv", header = TRUE, sep = "\t")
twoss <- read.delim(file = "../rtables/twoclass-auc-statsig.tsv", header = TRUE, sep = "\t")
threeauc <- read.delim(file = "../rtables/threeclass-auc.tsv", header = TRUE, sep = "\t")
threesig <- read.delim(file = "../rtables/threeclass-sig.tsv", header = TRUE, sep = "\t")
idcount <- read.delim(file = "../rtables/idcount.tsv", header = FALSE, sep = "\t")
phagecount <- read.delim(file = "../rtables/phagecount.tsv", header = FALSE, sep = "\t")
corstats <- read.delim(file = "../rtables/cor-stats.tsv", header = TRUE, sep = "\t")

pvalcalc <- function(x) {
  return(ifelse(x < 0.001, "< 0.001", paste("=", signif(x, digits = psigfig), sep = " ")))
}
```

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer). Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases suggest that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans have died from the disease [@Siegel:2014jo]. An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those changes might impact cancer progression and severity. We also aimed to evaluate the virome's potential for use as a diagnostic biomarker. The implications of this study are threefold. *First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Cohort Design, Sample Collection, and Processing
Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 who had carcinomas **(Figure \ref{sampleproc})**. Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The other half of each stool sample was purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active and extracellular virome* because we only sequenced those viruses that were encapsulated.

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction was performed with a blank buffer control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OVU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis. We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Similar to what was previously seen when analyzing the 16S rRNA gene sequences (from the same samples) [@Zackular:2014fba] and metagenomic samples [@Zeller:2014ix], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.

@@

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model yielded a statistically significant but minor performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that the combined features from the virus and bacterial communities improve our ability to classify stool as belonging to individuals with cancerous colons.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**. Although the features in the viral model were also of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven by a few OVUs, suggesting a role for the viruses in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome. This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Shifted Phage Influence Across Cancer Stages
We were able to evaluate shifts in influential phage identities during cancer progression because our cohort included people with healthy, adenomatous, and cancerous colons. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. Not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.

## Bacteriophage Dominance in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

@@

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network. We investigated the infectious capabilities and potential host ranges of all phage OVUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**). The predicted interactions were then used to identify community hub phages. This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OVU's connection to the rest of the network, and compared the centrality to the importance of each OVU in the colorectal cancer classification model. The phages with high centrality values were defined as community hubs. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

@@

A notable observation from our analysis was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences. This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing. We found that there were six bacterial OTUs that drove the performance of the 16S rRNA gene classification model, and these OTUs were all sparsely present and lowly abundant. Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model. The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application. There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but the previous approach has not utilize OGU clustering like we did here, it relied on lowly abundant signatures, and the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix]. Thus the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples. Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.

@@

## Analysis Source Code & Data Availability
All study sequences are available on the NCBI Sequence Read Archive under the following accession IDs:

All associated source code is available at the following GitHub repository:

@@

## Virus Metagenomic Library Preparation & Sequencing
Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez]. Briefly, an aliquot of stool (~0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension. The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22-µm filter to remove smaller contaminants. The filtered supernatant was treated with chloroform to lyse contaminating cells including bacteria, human, fungi, etc. The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase. The DNA was extracted from the purified VLPs using the Wizard PCR Purification Preparation Kit (Promega) [@Brum:2016jl]. Disease classes were staggered across purification runs to prevent run variation as a confounding factor. Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol. The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution. The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz]. The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
